---
title: 'Case Study 1: Predicting Algae Blooms'
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r}
data(algae, package="DMwR2")
algae
```

```{r}
summary(algae)
```
## Data Visualization

```{r}
hist(algae$mxPH, breaks = 30)
```
```{r}
library(Hmisc)
hist(subset(algae, select = c("mxPH", "mnO2", "Cl" ,"NO3", "NH4", "oPO4", "PO4", "Chla", "a1", "a2", "a3", "a4", "a5", "a6", "a7")), breaks = 20)
```

```{r}
boxplot(algae$oPO4, main = "Boxplot for orthophosphate")
```
```{r}
par(mfrow = c(1, 3))
boxplot(algae[algae$size == "small",]$a1, main  = "Small A1")
boxplot(algae[algae$size == "medium",]$a1, main  = "Medium A1")
boxplot(algae[algae$size == "large",]$a1, main  = "Large A1")
```
## Correlation matrix
```{r}
library(corrplot)
M = cor(subset(algae, select = c("mxPH", "mnO2", "Cl" ,"NO3", "NH4", "oPO4", "PO4", "Chla", "a1", "a2", "a3", "a4", "a5", "a6", "a7")))
corrplot(M, method = 'number',  tl.cex = 0.8, number.cex = 0.7)
```

## Checking for NAs in Data

```{r}
sum(is.na(algae))
```
## Columns with na
```{r}
which(colSums(is.na(algae))>0)
```
## Removing rows with na
```{r}
algae <- na.omit(algae)
```
```{r}
which(colSums(is.na(algae))>0)
```
## Correlation matrix
```{r}
M = cor(subset(algae, select = c("mxPH", "mnO2", "Cl" ,"NO3", "NH4", "oPO4", "PO4", "Chla", "a1", "a2", "a3", "a4", "a5", "a6", "a7")))
corrplot(M, method = 'number', tl.cex = 0.8, number.cex = 0.7)
```
```{r}
algae
```
# Regression for a1
```{r}
lm.a1 <- lm(a1 ~ ., data = algae[, 1:12])
summary(lm.a1)
```
```{r}
anova(lm.a1)
```
### Removing season from the model
```{r}
lm2.a1 <- update(lm.a1, . ~ . - season)
summary(lm2.a1)
```
```{r}
anova(lm.a1,lm2.a1)
```
```{r}
final.lm <- step(lm.a1)
```
```{r}
summary(final.lm)
```
## Regression Tree

```{r}
library(rpart)
rt.a1 <- rpart(a1 ~ ., data = algae[, 1:12])
rt.a1
```
```{r}
library(rpart.plot)
prp(rt.a1,extra=101,box.col="orange",split.box.col="grey")
```
```{r}
printcp(rt.a1)
```
### Pruning
```{r}
rt2.a1 <- prune(rt.a1, cp = 0.08)
rt2.a1
```
```{r}
lm.predictions.a1 <- predict(final.lm, algae)
rt.predictions.a1 <- predict(rt.a1, algae)
```
## Mean Absolute Error
```{r}
(mae.a1.lm <- mean(abs(lm.predictions.a1 - algae[["a1"]])))
(mae.a1.rt <- mean(abs(rt.predictions.a1 - algae[["a1"]])))
```
## Mean Squared Error
```{r}
(mse.a1.lm <- mean((lm.predictions.a1 - algae[["a1"]])^2))
(mse.a1.rt <- mean((rt.predictions.a1 - algae[["a1"]])^2))
```
## Normalized Mean Squared Error
```{r}
(nmse.a1.lm <- mean((lm.predictions.a1-algae[['a1']])^2)/mean((mean(algae[['a1']])-algae[['a1']])^2))
(nmse.a1.rt <- mean((rt.predictions.a1-algae[['a1']])^2)/mean((mean(algae[['a1']])-algae[['a1']])^2))
```

# Adding mean for nas
```{r}
data(algae, package="DMwR2")
algae
```
## Checking for NAs
```{r}
which(colSums(is.na(algae))>0)
```

```{r}
mean_val <- colMeans(algae[,4:18],na.rm = TRUE)
# replacing NA with mean value of each column
for(i in colnames(algae))
  algae[,i][is.na(algae[,i])] <- mean_val[i]
algae
```
```{r}
which(colSums(is.na(algae))>0)
```
## Correlation matrix
```{r}
M = cor(subset(algae, select = c("mxPH", "mnO2", "Cl" ,"NO3", "NH4", "oPO4", "PO4", "Chla", "a1", "a2", "a3", "a4", "a5", "a6", "a7")))
corrplot(M, method = 'number', tl.cex = 0.8, number.cex = 0.7)
```

## Removing PO4 because of having high correlation with oPO4
```{r}
algae = subset(algae, select = -PO4)
algae
```

```{r}
dim(algae)
```

```{r}
x = algae[,1:11]
ind<-sample(1:2, size=nrow(x), replace=TRUE, prob =c(0.7, 0.3))
train<-x[ind==1,]
test<-x[ind==2,]
```
```{r}
train
```
```{r}
test
```
## Regression on training data
```{r}
lm.a1 <- lm(a1 ~ ., data = train)
summary(lm.a1)
```
```{r}
anova(lm.a1)
```
```{r}
lm2.a1 <- update(lm.a1, . ~ . - season)
summary(lm2.a1)
```
## Predictions
```{r}
preds = predict(lm2.a1, test)
preds
```
```{r}
rmse_lm = sqrt(mean((test$a1 - preds)^2))
rmse_lm
```
## Regression Tree
```{r}
rt.a1 <- rpart(a1 ~ ., data = train)
rt.a1
```

```{r}
prp(rt.a1,extra=101,box.col="orange",split.box.col="grey")
```

```{r}
rt2.a1 <- prune(rt.a1, cp = 0.08) 
rt2.a1
```
## Prediction on Regression Tree
```{r}
rt_preds = predict(rt2.a1, test)
rt_preds
```
```{r}
rmse_rt = sqrt(mean((test$a1 - rt_preds)^2))
rmse_rt
```










